Towards Proof-Reconstruction of Problems in Classical Propositional Logic

### 2017-10-02 paper/sections/proof-reconstruction.tex
[ paper ] making some notes.

With such a derivation, in Step~5, we process the derivation with our \Haskell translator tool, \Athena~\cite{Athena}.
\Athena parses the \TSTP format, analyze the
derivation and generates a representation of the natural deduction
proof using a tree data structure (see the properties
of this tree in Subsection \ref{ssec:metis-proofs}).
We perform an analysis of the proof-tree removing redundancies,
unnecessary steps or unused output from the \TSTP derivation.

@@

The definition above suffers termination problems since it is not
a structural recursion. Therefore, we have reformulated Def.~\ref{eq:split-definition} to solve such a issue following the technique
described in Subsection~\ref{ssec:structural-recursion}.
Here and below, we refer to $split$ as the function after applying
the process mentioned above.

@@

Syntactically speaking, showing the logic equivalence between two conjunctive normal forms is a matter of rearrangements of the conjunctions
but also of the inner disjunctions inside the propositions involved.

@@

The following function, $build_{∨} : \Prop → \Prop → \Prop$ in Eq.~\ref{eq:build-or}, intends to construct a disjunction from another disjunction, specifically, we want to rearrange the order of the disjuncts in a target disjunction.
This approach for rearrangement is inspired on the treatment for nested
disjunctions mainly described in \citeauthor{bohme2010}~\cite{bohme2010}.

@@

Now, with a right-associative proposition,
$sbuild_{∨}: \Prop → \Prop → \Prop$ in
Eq.~\ref{eq:strong-build-or-definition} aims to construct the disjunction
ψ by using each of disjuncts from the proposition φ. This approach to
build a disjunction from another is inspired on the description
\citeauthor{bohme2010}~\cite{bohme2010} to prove equality between nested
disjunctions.

@@

\begin{lemma}[lem-conjunct$_{∨}$]
  \label{lem:lem-conjunct-or}
  $Γ ⊢ φ$, and $ψ \in\Prop$ then $Γ ⊢ conjunct_{∨}(φ, ψ)$, where,
  \begin{align*}
  \begin{split}
    \label{eq:conjunct-or-definition}
    conjunct_{∨}(φ, ψ) =
    \begin{cases}
      ψ, &\text{if }reorder_{∨}(φ, ψ) ≡ ψ\\
      ψ, &\text{if }ψ ≡ ψ₁ ∧ ψ₂, conjunct_{∨}(φ, ψ₁) ≡ ψ₁\\
        &\text{and } reorder_{∨}(φ, ψ₂) ≡ ψ₂\\
      ψ, &\text{if }φ ≡ φ₁ ∧ φ₂, conjunct_{∨}(φ₁, ψ) ≡ ψ\\
      ψ, &\text{if }φ ≡ φ₁ ∧ φ₂, conjunct_{∨}(φ₂, ψ) ≡ ψ\\
      φ, &\text{otherwise.}
    \end{cases}
  \end{split}
  \end{align*}
\end{lemma}

@@

The \resolve rule is a generalization version for resolution in propositional logic since the formulas are not shaped always as the theorem expect (see Fig.~\ref{fig:metis-inferences}).

@@

\subsubsection{Normal Forms.}
\label{sssec:normal_forms}

A negative normal form for the formula φ is one in which negations
appear only in the literals and the expression is in terms only of
the boolean operators (¬, ∧, ∨).

@@

\begin{lemma}[lem-dist$_{∨}$]
  \label{lem:lem-dist-or}
  $Γ ⊢ φ$ then $Γ ⊢ dist_{∨}(φ)$, where,
  \begin{align*}
      \begin{split}
      dist_{∨} =
        \begin{cases}
          dist_{∨}(φ₁ ∨ φ₂) ∧ dist_{∨}(φ₂ ∨ φ₃),
            &\text{if }φ ≡ (φ₁ ∧ φ₂) ∨ φ₃\\
          dist_{∨}(φ₁ ∨ φ₃) ∧ dist_{∨}(φ₁ ∨ φ₃),
            &\text{if }φ ≡ φ₁ ∨ (φ₂ ∧ φ₃)\\
          φ, &\text{otherwise.}
        \end{cases}
      \end{split}
  \end{align*}
\end{lemma}

\begin{lemma}[lem-dist]
  \label{lem:lem-dist}
  $Γ ⊢ φ$ then $Γ ⊢ dist(φ)$, where,
  \begin{align*}
      \begin{split}
      dist =
        \begin{cases}
          dist(φ₁) ∧ dist(φ₂),
            &\text{if }φ ≡ φ₁ ∧ φ₂\\
          dist_{∨}(dist(φ₁) ∨ dist(φ₂)),
            &\text{if }φ ≡ φ₁ ∨ φ₂\\
          φ, &\text{otherwise.}
        \end{cases}
      \end{split}
  \end{align*}
\end{lemma}

\begin{proof} Sketch. Case when $φ ≡ φ₁ ∨ φ₂$.\\
For $i = 1, 2$,
\begin{equation}
  \begin{bprooftree}
    \RightLabel{assume}
    \AxiomC{$Γ, φ_{i} ⊢ φ_{i}$}
    \RightLabel{lem-dist}
    \UnaryInfC{$Γ ⊢ dist(φ_{i})$}
    \RightLabel{∨-intro$_{1,2}$}
    \UnaryInfC{$Γ, φ_{i} ⊢ dist(φ₁) ∨ dist(φ₂)$}
    \RightLabel{lem-dist$_{∨}$}
    \UnaryInfC{$Γ, φ_{i} ⊢ dist_{∨}(dist(φ₁) ∨ dist(φ₂))$.}
    \end{bprooftree}
\end{equation}
Then, by using the ∨-elim rule, from the last derivation above we derive
$Γ, φ₁ ∨ φ₂ ⊢ dist_{∨}(dist(φ₁) ∨ dist(φ₂))$, and the lemma follows.
\end{proof}

@@

@@

\begin{lemma}[lem-canon]
  \label{lem:lem-canon}
  \begin{align*}
  \begin{split}
  &canon(φ) =\\
  &\hspace{3mm}
  \begin{cases}
      canon(φⱼ), &\text{if } φ ≡ φ₁ ∧ φ₂,\ canon(φᵢ) ≡ ⊤, i,j = 1,2, i≠j\\
     ⊥, &\text{if } φ ≡ φ₁ ∧ φ₂,\ canon(φᵢ) ≡ ⊥, i = 1,2\\
     canon(φ₁) ∧ canon(φ₂), &\text{if } φ ≡ φ₁ ∧ φ₂\\
     ⊤, &\text{if } φ ≡ φ₁ ∨ φ₂,\ canon(φᵢ) ≡ ⊤, i = 1,2\\
     canon(φⱼ), &\text{if } φ ≡ φ₁ ∨ φ₂,\ canon(φᵢ) ≡ ⊥,  i,j = 1,2, i≠j\\
     canon(φ₁) ∨ canon(φ₂), &\text{if } φ ≡ φ₁ ∨ φ₂\\
     φ &\text{otherwise.}
    \end{cases}
  \end{split}
  \end{align*}
\end{lemma}

@@

>>>>

With such a derivation, in Step~5, we process the derivation with our \Haskell translator tool, \Athena~\cite{Athena}.
\Athena parses the \TSTP format, analyze the
derivation and generates a representation of the natural deduction
proof using a tree data structure (see the properties
of this tree in Section~\ref{ssec:metis-proofs}).
We perform an analysis of the proof-tree removing redundancies,
unnecessary steps or unused output from the \TSTP derivation.

@@

The definition above suffers termination problems since it is not
a structural recursion. Therefore, we have reformulated Def.~\ref{eq:split-definition} to solve such a issue following the technique
described in Section~\ref{ssec:structural-recursion}.
Here and below, we refer to $split$ as the function after applying
the process mentioned above.

@@

Syntactically speaking, showing the logic equivalence between two conjunctive normal forms is a matter of rearrangements of the conjunctions
but also of the inner disjunctions inside the propositions involved.
\improvement{Make clear: we are dealing with CNFs, and prove equivalence between two formulas is equivalent to show two trees are  isomorphic.
And that problem is NP-hard, and for that, we mostly do brute force.}

@@

The following function, $build_{∨} : \Prop → \Prop → \Prop$ in Eq.~\ref{eq:build-or}, intends to construct a disjunction from another disjunction, specifically, we want to rearrange the disjuncts in the source formula to match with the target disjunction for syntactic
equivalence.

@@

Now, with a right-associative disjunction by applying $rdisj$ function to
a proposition $φ$, $sbuild_{∨}: \Prop → \Prop → \Prop$ in
Eq.~\ref{eq:strong-build-or-definition} aims to construct the disjunction
ψ by using each of its disjuncts from the proposition φ. This approach to
build a disjunction from another is inspired on the description in
\citeauthor{bohme2010}~\cite{bohme2010} to prove equality between nested
disjunctions. \change{Make clear this, It's not clear the difference with build-or and sbuild-or.}

@@

\begin{lemma}[lem-conjunct$_{∨}$]
  \label{lem:lem-conjunct-or}
  $Γ ⊢ φ$, and $ψ \in\Prop$ then $Γ ⊢ conjunct_{∨}(φ, ψ)$, where,
  \improvement{The first case is not implemented in Agda}
  \begin{align*}
  \begin{split}
    \label{eq:conjunct-or-definition}
    conjunct_{∨}(φ, ψ) =
    \begin{cases}
      ψ, &\text{if }φ ≡ ψ\\
      ψ, &\text{if }reorder_{∨}(φ, ψ) ≡ ψ\\
      ψ, &\text{if }ψ ≡ ψ₁ ∧ ψ₂, conjunct_{∨}(φ, ψ₁) ≡ ψ₁\\
        &\text{and } reorder_{∨}(φ, ψ₂) ≡ ψ₂\\
      ψ, &\text{if }φ ≡ φ₁ ∧ φ₂, conjunct_{∨}(φ₁, ψ) ≡ ψ\\
      ψ, &\text{if }φ ≡ φ₁ ∧ φ₂, conjunct_{∨}(φ₂, ψ) ≡ ψ\\
      φ, &\text{otherwise.}
    \end{cases}
  \end{split}
  \end{align*}
\end{lemma}

@@

The \resolve rule is a generalization version for resolution in propositional logic since the formulas are not shaped always as the aforementioned theorem expects (see Fig.~\ref{fig:metis-inferences}).

@@

\subsubsection{Normal Forms.}
\label{sssec:normal_forms}
\change{Here lacks of an introduction!}
A negative normal form for the formula φ is one in which negations
appear only in the literals and the expression is in terms only of
the boolean operators (¬, ∧, ∨).

@@

\begin{lemma}[lem-dist$_{∨}$]
  \label{lem:lem-dist-or}
  $Γ ⊢ φ$ then $Γ ⊢ dist_{∨}(φ)$, where,
  \begin{align*}
      \begin{split}
      dist_{∨}(φ) =
        \begin{cases}
          dist_{∨}(φ₁ ∨ φ₂) ∧ dist_{∨}(φ₂ ∨ φ₃),
            &\text{if }φ ≡ (φ₁ ∧ φ₂) ∨ φ₃\\
          dist_{∨}(φ₁ ∨ φ₃) ∧ dist_{∨}(φ₁ ∨ φ₃),
            &\text{if }φ ≡ φ₁ ∨ (φ₂ ∧ φ₃)\\
          φ, &\text{otherwise.}
        \end{cases}
      \end{split}
  \end{align*}
\end{lemma}

\begin{lemma}[lem-dist]
  \label{lem:lem-dist}
  $Γ ⊢ φ$ then $Γ ⊢ dist(φ)$, where,
  \begin{align*}
      \begin{split}
      dist(φ) =
        \begin{cases}
          dist(φ₁) ∧ dist(φ₂),
            &\text{if }φ ≡ φ₁ ∧ φ₂\\
          dist_{∨}(dist(φ₁) ∨ dist(φ₂)),
            &\text{if }φ ≡ φ₁ ∨ φ₂\\
          φ, &\text{otherwise.}
        \end{cases}
      \end{split}
  \end{align*}
\end{lemma}

\begin{sketchproof}
Case when $φ ≡ φ₁ ∨ φ₂$.\\
For $i = 1, 2$,
\begin{equation}
  \begin{bprooftree}
    \RightLabel{assume}
    \AxiomC{$Γ, φ_{i} ⊢ φ_{i}$}
    \RightLabel{lem-dist}
    \UnaryInfC{$Γ ⊢ dist(φ_{i})$}
    \RightLabel{∨-intro$_{1,2}$}
    \UnaryInfC{$Γ, φ_{i} ⊢ dist(φ₁) ∨ dist(φ₂)$}
    \RightLabel{lem-dist$_{∨}$}
    \UnaryInfC{$Γ, φ_{i} ⊢ dist_{∨}(dist(φ₁) ∨ dist(φ₂))$.}
    \end{bprooftree}
\end{equation}
Then, by using the ∨-elim rule, from the last derivation above we derive
$Γ, φ₁ ∨ φ₂ ⊢ dist_{∨}(dist(φ₁) ∨ dist(φ₂))$, and the lemma follows.
\end{sketchproof}

@@

\improvement{We need here a text, or a comment about what
is going next.}

@@

\begin{lemma}[lem-canon]
  \label{lem:lem-canon}
  \begin{align*}
  \begin{split}
  &canon(φ) =\\
  &\hspace{3mm}
  \begin{cases}
      canon(φⱼ), &\text{if } φ ≡ φ₁ ∧ φ₂,\ canon(φᵢ) ≡ ⊤,\\
                 &\text{for same } i,j = 1,2, i≠j\\
     ⊥, &\text{if } φ ≡ φ₁ ∧ φ₂,\ canon(φᵢ) ≡ ⊥, i = 1,2\\
     canon(φ₁) ∧ canon(φ₂), &\text{if } φ ≡ φ₁ ∧ φ₂\\
     ⊤, &\text{if } φ ≡ φ₁ ∨ φ₂,\ canon(φᵢ) ≡ ⊤, i = 1,2\\
     canon(φⱼ), &\text{if } φ ≡ φ₁ ∨ φ₂,\ canon(φᵢ) ≡ ⊥,\\
                &\text{for same } i,j = 1,2, i≠j\\
     canon(φ₁) ∨ canon(φ₂), &\text{if } φ ≡ φ₁ ∨ φ₂\\
     φ, &\text{otherwise.}
    \end{cases}
  \end{split}
  \end{align*}
\end{lemma}

@@

The \simplify rule is the application of the the lemma showed above,
but also the following lemmas that intend to reduce a list of derivations,
a conjunction of their formulas into a one smaller.

\improvement{Missing theorems here.}
